---
title: "Class 19:"
author: "Barry (PID 911)"
format: html
---

# Web scrapping

Here I extract the CDC figures for Pertussis cases in USA:
https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html


```{r echo=FALSE}
cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L),
         Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617)
       )
```

Let's make a plot of number of Pertussis cases per year.

```{r}
library(ggplot2)

base <- ggplot(cdc) +
  aes(Year, Cases) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels = scales::label_comma())

base
```

The first vaccine (so called wP) was introduced in the USA in 1946. Let's add this ass a line to our plot 

```{r}
base + geom_vline(xintercept = 1946, col="blue")
```


The US, and many other countries, switched from the original wP vaccine to a new aP vaccine (acellular) in 1996. Let's add this to our plot.

```{r}
base + geom_vline(xintercept = 1946, col="blue") +
  geom_vline(xintercept = 1996, col="red")

```


# Exploring CMI-PB data

Why is this vaccine-preventable disease on the upswing? To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

The new and ongoing **CMI-PB** project aims to provide the scientific community with this very information.

The CMI-PB resource API returns JSON format. We will use the **jsonlite**.

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/subject",
                     simplifyVector = TRUE)

head(subject)
```

> Q. How many wP and aP subject there are?

```{r}
table(subject$infancy_vac)
```


> Q. How many female non-white individuals are there in the dataset.

```{r}
table(subject$race, subject$biological_sex)
```

Let's look at the specimen table next.

```{r}
specimen <- read_json("http://cmi-pb.org/api/specimen",
                      simplifyVector = TRUE)

head(specimen)
dim(specimen)
```


To know whether a given specimen_id comes from an aP or wP individual we need to link (a.k.a. “join” or merge) our specimen and subject data frames. The excellent dplyr package (that we have used previously) has a family of join() functions that can help us with this common task:


```{r}
library(dplyr)
meta <- inner_join(specimen, subject)
dim(meta)
```

Read one of the main expermental data tables `ab_titer`

```{r}
titer <- read_json("http://cmi-pb.org/api/ab_titer",
                   simplifyVector = TRUE)

dim(titer)
head(titer)
```

> Q. How many isotypes do we have in this dataset?

```{r}
table(titer$isotype)
```

Do see other info we need to join titer and meta tables to put all the data back together.


```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```
```{r}
head(abdata)
```

> Q. What do you notice about the "vist" number 8?

```{r}
table(abdata$visit)
```

# Examine IgG1 Ab titer levels

```{r}
ig1 <- filter(abdata, isotype=="IgG1", visit!=8)
dim(ig1)
head(ig1)
```

> Q. How many antigens are there

```{r}
table(abdata$antigen)
```

Analysis of the whole dataset: antigen levels - i.e. a plot of antigen vs MFI...

```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot()
```

Add faceting by visit.


```{r}
ggplot(ig1) +
  aes(MFI, antigen) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

Looks like all the antigens that change over time are in the vaccine!
PT, TT, PRN, FIM2/3 etc.

What about wP vs aP? Are there any obvious differences?

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))

```

or make a version where we facet with aP/wP

```{r}
ggplot(ig1) +
  aes(MFI, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(infancy_vac, visit), nrow=2) +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
```


> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“Measles”, that is not in our vaccines) and a clear antigen of interest (“FIM2/3”, extra-cellular fimbriae proteins from B. pertussis that participate in substrate attachment).


```{r}
filter(ig1, antigen=="Measles") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  labs(title="Measles antigen levels") +
  theme_bw()
```

and for FIM2/3

```{r}
filter(ig1, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  labs(title="FIM2/3 antigen levels") +
  theme_bw()
```

```{r}
filter(ig1, antigen=="PT") %>%
  ggplot() +
  aes(MFI, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  labs(title="PT antigen levels") +
  theme_bw()
```

> Q17. Do you see any clear difference in aP vs. wP responses?

No not really...

# Obtaining CMI-PB RNASeq data

For RNA-Seq data the API query mechanism quickly hits the web browser interface limit for file size. We will present alternative download mechanisms for larger CMI-PB datasets in the next section. However, we can still do “targeted” RNA-Seq querys via the web accessible API.

For example we can obtain RNA-Seq results for a specific ENSEMBLE gene identifier or multiple identifiers combined with the & character:


```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 

head(rna)
```

We need another `inner_join()` here to get all the metadata

```{r}
ssrna <- inner_join(rna, meta)
```
```{r}
head(ssrna)
```

> Q18. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).


```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```




