---
title: "Class 5: Data Viz with ggplot"
author: "Barry (PID: 911)"
format: gfm
---

R has lot's of ways to make figures and graphs in particular.
One that comes with R out of the box is called **"base" R** - the `plot()` function.

```{r}
plot(cars)
```

A very popular package in this area is called **ggplot2**.

Before I can use any add-on package like this I must install it with the `install.packages("ggplot2")` command/function.

Then to use the package I need to load it with a `library(ggplot2)` call.


```{r}
library(ggplot2)

ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

For "simple" plots like this one base R code will be much shorter than ggplot code.

Let's fit a model and show it on my plot:

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth() 
```

Every ggplot has at least 3 layers

- **data** (data.frame with the numbers and stuff you want to plot)
- **aes**thetics (mapping of your data columns to your plot)
- **geom**s (there are tones of these, basics are `geom_point()`, `geom_line()`, `geom_col()`)

```{r}
head(mtcars)
```

Make me a ggplot of the `mtcars` data set using `mpg` vs `disp`
and set the size of the points to the `hp`
and set the color to `am`

```{r}
ggplot(mtcars) +
  aes(x=mpg, y=disp, size=hp, col=am) +
  geom_point()
```

## Gene expression plot

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
nrow(genes)
```

There are `r nrow(genes)` genes in this dataset.

```{r}
unique(genes$State)
```
The `table()` function is a super useful utility to tell me how many entries of each type there are.

```{r}
round( table(genes$State) / nrow(genes), 3)
```

The functions `nrow()`, `ncol()`, and `table()` are ones I want you to know.

> Key points:
Saving plots with **ggsave()**
Different plot "types" with different `geoms_**()`
Faceting with `facet_wrap()`
Multi-plot layout with the **patchwork** package.

```{r}
p <- ggplot(mtcars) +
  aes(mpg, disp) +
  geom_point()

#ggsave("myplot.pdf")
```



```{r}
p
```

