---
title: "Class 5: Data Visualization"
author: "Barry"
format: pdf
---

# Plotting in R

R has many plotting and visualization systems including "base" R.

```{r}
head(cars, n=6)
```

## A base plot

```{r firstplot}
plot(cars)
```

Base R plots can be quite simple for basic plots when compared to systems like ggplot.

To use an add on package, like ggplot, I have to first get it on my computer - i.e. install it!

We use the function `install.packages()` with the name of the package we want to install.

```{r}
library(ggplot2)
ggplot(cars)
```

ggplot is much more verbose than base R plot and every single ggplot needs at least 3 things:

-   **Data** (this is the data.frame with the stuff we want to plot)
-   **Aesthetics** or aes() for short (how the data map to the plot)
-   **Geoms** (like geom_point(), geom_line() the plot type)

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  theme_bw()
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(se=FALSE, method="lm")
```

## A plot of some gene expresion data

The code to read the data:

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"

genes <- read.delim(url)

head(genes)
```


> Q. How many genes are in this dataset?

```{r}
nrow(genes)
```

There are `r nrow(genes)` genes in this dataset.

> How many genes are up-regulated?

```{r}
table( genes$State )
```


```{r}
sum(genes$State == "up")
```

Let's stop messing and plot it

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, color=State) +
  geom_point()
```

I can save any ggplot object for use later so I don't need to type it all out again. Here I save my starting plot to the object `p` then I can add layers to `p` later on.


```{r}
p <- ggplot(genes) + 
    aes(x=Condition1, y=Condition2, col=State) +
    geom_point()
```

```{r}
p + scale_colour_manual( values=c("blue","gray","red") ) +
  labs(title="Gene expresion changes upon drug treatment",
       subtitle = "Some subtitle")
```
My object `p` is just the same.

```{r}
p
```


```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, color=State) +
  geom_point(shape=21)
```

## A more complex ggplot example 

One of the big wins with ggplot is how easy it is to facet your data into sub-plots..

Read the gapminder dataset from online

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```


and have a wee peak

```{r}
head(gapminder)
```
> Q. How many countrys are in this dataset?

```{r}
length( unique(gapminder$country) )
```

> Q. How many years do we have data for?

```{r}
min(gapminder$year)
max(gapminder$year)
```

```{r}
range(gapminder$year)
```

> Q. Which country has the smallest population in the dataset?

```{r}
min(gapminder$pop)
```
First where is this min value in the pop vector

```{r}
ind <- which.min(gapminder$pop)
```

Now use this to access the $country value for this position

```{r}
gapminder$country[ind]
```

```{r}
gapminder[ind,]
```

Make a first plot of gdb vs life exp

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, color=continent) +
  geom_point(alpha=0.7) +
  facet_wrap(~continent)
```

