---
title: "Class17: Vaccination Rate Mini Project"
author: "Barry (PID 911)"
format: gfm
---

# Background

In today's class we will explore a data-set on state wide vaccination rate from CA.Gov

Our goal is to examine and compare the Covid-19 vaccination rates around San Diego.

We will start by downloading the most recently dated “Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code


## Data Input

We can use read.csv() to import our data into R

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

# Quick EDA

> Q1. What column details the total number of people fully vaccinated?

```{r}
head(vax$persons_fully_vaccinated)
```


> Q2. What column details the Zip code tabulation area?

```{r}
head(vax$zip_code_tabulation_area)
```

> Q3. What is the earliest date in this dataset?

```{r}
vax$as_of_date[1]
```
> Q4. What is the last date

```{r}
vax$as_of_date[ nrow(vax) ]
```

A useful function for exploring new datasets is from the **skimr** package:

```{r}
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset?

13

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

```{r}
n.missing <- sum(is.na(vax$persons_fully_vaccinated))
n.missing
```

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

```{r}
round( n.missing/nrow(vax) * 100, 2)
```

# Working with dates

We will use the **lubridate** package to help ease the pain of working with times and dates.

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
 today() - ymd(vax$as_of_date[1])
```

```{r}
today() - ymd("1999-4-14")
```

Sad news

```{r}
today() - ymd("1978-6-22")
```

> Q9. How many days have passed since the last update of the dataset?

I will convert the entire 'as_of_date' column to be in lubridate format.

```{r}
# Specify that we are using the year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
today() - vax$as_of_date[ nrow(vax) ]
```


> Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length( unique(vax$as_of_date) )
```

```{r}
library(dplyr)

n_distinct(vax$as_of_date)
```

# Working with ZIP codes

There are quite a few R packages that can help ease the pain of working with ZIP codes. We will try just one of the smaller ones here **zipcodeR**

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```
```{r}
geocode_zip("48044")
```

Calculate the distance between the centroids of any two ZIP codes in miles, e.g.

```{r}
zip_distance('92037','92109')
```

```{r}
reverse_zipcode(c('92037', "92109") )
```

> Q. Find the best and worst ratio of "median household income" to "median_home_value" in San Diego county.

First find all San Diego county ZIP codes.

```{r}
sd.zip <- unique( vax$zip_code_tabulation_area[vax$county == "San Diego"])
length(sd.zip)
```
Now do a zipcode lookup for the data we want

```{r}
sd.eco <- reverse_zipcode( sd.zip )
```

Now extract the "median household income" and "median_home_value" to help answer our question above.

Most expensive area?
```{r}
ord <- order(sd.eco$median_home_value, decreasing=TRUE)
head( sd.eco[ord,] )
```

```{r}
head( arrange(sd.eco, desc(median_home_value)) )
```

# Focus on the San Diego area

I will use **dplyr** here to help me do more involved selections (i.e. filter rows to include the subset of data we are intrested in).

```{r}
sd <- filter(vax, county == "San Diego")
nrow(sd)
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
n_distinct(sd$zip_code_tabulation_area)
```

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
ind <- which.max(sd$age12_plus_population)
sd[ind,]
```

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2023-03-07”?

```{r}
vax$as_of_date[ nrow(vax) ]
```

```{r}
sd.latest <- filter(sd, as_of_date == "2023-03-07")
mean( sd.latest$percent_of_population_fully_vaccinated, na.rm=TRUE )
```

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2023-02-28”?


```{r}
hist(sd.latest$percent_of_population_fully_vaccinated)
```

```{r}
library(ggplot2)

ggplot(sd.latest) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(bins=20) +
  theme_bw() +
  xlab("Precent of People Fully Vaccinated")
```



# Focus on UCSD/La Jolla

UC San Diego resides in the 92037 ZIP code area and is listed with an age 5+ population size of 36,144

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```
Make a time course plot of vaccination progress for this zip code area

```{r}
ljplot <- ggplot(ucsd) +
  aes(x=as_of_date, y=percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line() +
  xlab("Date") +
  ylab("Percent of Population Fully Vaxed") +
  labs(title = "Vaccination Rate for La Jolla (CA 92037)") +
  theme_bw()

ljplot
```

### Comparing to similar sized areas

 Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on as_of_date “2023-03-07”
 
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2023-03-07")

head(vax.36)
```

How many ZIP code areas are we talking about here?

```{r}
n_distinct(vax.36$zip_code_tabulation_area)
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2023-03-07”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
ave.36 <- mean( vax.36$percent_of_population_fully_vaccinated, na.rm=TRUE )
ave.36
```

```{r}
ljplot + geom_hline(yintercept = ave.36, col="red", linetype=2)
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2023-03-07”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

```{r}
ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() +
  xlim(0,1)
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
filter(vax.36, zip_code_tabulation_area %in% c("92109", "92040"))
```

```{r}
val <- vax %>% filter(as_of_date == "2023-03-07") %>%  
  filter(zip_code_tabulation_area %in% c("92109", "92040")) %>%
  select(percent_of_population_fully_vaccinated)

val
```

```{r}
val < ave.36
```


Both are below average!


```{r}
ggplot(vax.36) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram() +
  xlim(0,1) +
  geom_vline(xintercept = c(0.55, 0.69), col="blue", linetype=2)
```


> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.


```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)

ggplot(vax.36.all) +
  aes(x=as_of_date,
      y=percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  geom_hline(yintercept = ave.36) +
  labs(title = "Vaccination Rate accross Califorina",
       subtitle = "Only areas with a population above 36k are shown",
       x="Date",
       y="Percent of Population Fully Vaxed") +
  theme_bw()
```

