---
title: "Class 10: Candy Project"
author: "Barry"
format: html
---

Here we take 538's candy dataset and explore it with methods we have been learning in class (Hint, hint, PCA!).

# Import the data

```{r}
candy = read.csv("candy-data.csv", row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

Look up how my candy is doing?

An overview of the dataset
```{r}
library("skimr")
skim(candy)
```


> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent)
```
```{r}
win.choc <- candy[as.logical(candy$chocolate),"winpercent"]
win.fruit <- candy[as.logical(candy$fruity),"winpercent"]
```

```{r}
mean(win.choc)
mean(win.fruit)
```
```{r}
t.test(win.choc, win.fruit)
```

> Q13. What are the five least liked candy types in this set?

```{r}
sort( c(5,4,1,2) )
```
```{r}
order( c(5,4,1,2) )
```
```{r}
ord.ind <- order(candy$winpercent)
head(candy[ord.ind,] )
```


> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
ord.ind <- order(candy$winpercent, decreasing = TRUE)
head(candy[ord.ind,], 5 )

```

> Q15. Make a first barplot of candy ranking based on winpercent values.


```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col() + 
  theme_bw()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?


```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col() + 
  theme_bw()


```


Let's add some color. First we will make a color vector.


```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) + 
  theme_bw()


```

Now, for the first time, using this plot we can answer questions like:

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst


# 4. Taking a look at pricepercent

What about value for money? What is the the best candy for the least money? One way to get at this would be to make a plot of winpercent vs the pricepercent variable. 

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text(col=my_cols)
```

Let's change the pink to red so we can see it better

```{r}
my_cols[as.logical(candy$fruity)] = "red"
```



```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```


# 5 Exploring the correlation structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the `corrplot` package to plot a correlation matrix.

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

# Principal Component Analysis

Let’s apply PCA using the `prcom()` function to our candy dataset remembering to set the scale=TRUE argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Now we can plot our main PCA score plot of PC1 vs PC2.

```{r}
plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)
```

